\mythesistitleNOTSYNPOSIS

### 2013-03-23 src/response_to_reviewers.tex

\documentclass[12pt,a4paper]{article}
\usepackage{fullpage}
\usepackage[normalem]{ulem}
\usepackage{../lib/mythesislib}
\usepackage{../lib/mylatexlib}
\usepackage{setspace}
%

\begin{document}

@@

\maketitle

\onehalfspacing

\section{Reviewers}

\begin{enumerate}
\item  {\bf Dr. C. R. Subramaniam}\\ Institute of Mathematical Sciences, Chennai
\item  {\bf Dr. Subhas Chandra Nandy}\\ Indian Statistical Institute, Kolkata
\end{enumerate}
The reviewers careful comments are acknowledged and
appreciated. Responses to the same are given in following sections.

@@

\item Comment: {\bf In thesis Page 28, Definition 7 (of binary
    operator /): it is not defined precisely. What do you mean by
    ``$b$ is a new element added..''?}.\\
  Response: This was in thesis Page 26. $A/B$ is created by removing
  the elements of second operand set $B$ from the first operand set
  $A$ and introducing a new element {\em not} in the universe to $A$
  to represent set $B$. This is a set operation mentioned in
  \cite{mm96} (they use ``$B$'' to denote the new element to represent
  set $B$ which is also confusing) which is eventually used to
  decompose permutation of elements in $A$ to permutation of elements
  in $A/B$ and permutation of elements in $B$.  Perhaps it would be
  clearer if $b$ is replaced by $x_B$. This change has been made in
  Page 26 Definition 2.2.4 (7) and other mentions of this operation
  ``/'' Page 27 item (iii) and $h$ has been replaced with $x_H$ in
  Corollary 2.2.8. There was also a typo in Page 27 paragraph under
  Corollary 2.2.8 item (ii) where ``$U \setminus H$'' was written
  instead of $U / H$.  This has also been corrected.

\item Comment: {\bf In thesis Page 39 Definition 2.3.1: the definition
    of $A_\cF$ is not clear. Where is the dependence on $\cF$?}\\
  Response: This was in thesis Page 36. It is true, there is no
  dependence in $\cF$. It is only dependent on the universe $U$ of
  $\cF$. All occurences of $A_\cF$ has been changed to $A_U$ in Page
  36.
\end{enumerate}

\subsection{Other comments}
The following are a few observations mentioned in the review document
with my responses/changes.

\begin{enumerate}
\item Observation: (in Page 1, second last paragraph) {\bf The result
    on characterization of TPL in the thesis is a variant of the 1978
    JCT(B) result due to Fournier who showed that two hypergraphs $H_1
    = (V_1, X)$ and $H_2 = (V_2, Y)$ such that $|X| = |Y| = m$ are
    isomorphic iff there is a bijection $\phi$ on $I = \{1, ... , m\}$
    s.t. for every $L \subset I,
    |\bigcap_{i \in L}X_i| = |\bigcap_{j=\phi(i),i \in L}Y_j|$} \\
  Response: This is an important observation and was not cited in the
  thesis. Also \cite{f80} generalizes \cite{br72} and \cite{fg65} by
  characterizing the isomorphism of two hypergraphs by means of
  equicardinality of certain edge intersections and the exclusion of
  certain pairs of subhypergraphs. TPL characterization is a special
  case of this with one of the hypergraphs having hyperedges that are
  paths from a tree and the characterization only uses edge
  intersections of at most 3 hyperedges. This has been added in the
  Conclusions chapter Page 72 under {\em Graph isomorphism and
    logspace canonization}.

\item Observation: (in Page 2, first paragraph) {\bf A $k$-subdivided
    \sout{tree} [star] is a tree with a single central vertex with a number of paths
    each of $k$ edges emanating from it.}\\
  Response: $k$-subdivided star as defined in this thesis (Section 3.2.4,
  Page 43) is such that each of the aforementioned paths are of length
  ``$k+1$'' edges since each such path has $k$ nodes apart from the
  central node and the leaf node, i.e.  $k+2$ nodes. No change in the
  thesis.
\end{enumerate}

@@

\item Comment: {\bf The proof of Lemma 3.3.4 is unclear.}\\
Response: The proof was correct but terse. It has been elaborated to as
follows:

  {\em 
  Lemma:   Let $(\cF, \cl)$ be an \ICPPL and $\cl(S_i)=P_i$, $S_i \in \cF$, $1
  \leq i \leq 4$.  Then, $|\cap_{i=1}^4 S_i| = |\cap_{i=1}^4 P_i|$.

  Proof: Consider the set of set intersections with $S_1$ $\cF' = \{S_2 \cap S_1, S_3
  \cap S_1, S_4 \cap S_1\}$ and let $\cl'$ be a tree path labeling of
  $\cF'$ such that $\cl'(S_i \cap S_1) = P_i \cap P_1, 2 \le i \le
  4$. Clearly, $P_2 \cap P_1$, $P_3 \cap P_1$, and $P_4 \cap P_1$ are
  subpaths of path $P_1$, thus equivalent to intervals.  Due to the
  three way intersection cardinality preservation property of the
  ICPPL $(\cF, \cl)$, this new tree path labeling $(\cF', \cl')$
  preserves pairwise intersection cardinalities.  Now by applying
  Lemma 3.3.1 %~\ref{lem:intersection-cardinality} 
  to the sets in $\cF'$ and
  their corresponding path images due to $\cl'$, it follows that
  $|\cap_{i=1}^4 S_i| =| \cap_{i=1}^4 P_i|$.}

\item Comment: {\bf While studying Feasible Tree Path Labeling Problem
  in Section 3.3, two filters are mentioned and their pseudocodes
  given. If possible, state the problem in words. ... Explicitly state
the roles of these two filters in the algorithm with proper
figures.}\\
Response: The problems are described in words but they were part of a
long narrative. It has been summarised now in Page 44 before going
into details of the algorithms as follows:

{\em \begin{enumerate}
\item [Filter 1.] \filteri Refine $\cF$ such that the resulting
  labeling will not have paths that share a leaf thus each leaf being
  unique to a path. This is done by breaking the paths into subpaths and their
  corresponding preimage sets as described in Algorithm 1.
\item [Filter 2.] \filterii Find the element in universe $U$ that maps to each
  leaf in $T$ as described in Algorithm 2. 
\end{enumerate}
Remove the leaves from $T$ and their corresponding preimages from $U$
and call the filters again. This is repeated until the resulting
truncated tree is a path. The remaining mapping can be found using
ICPIA.
}

Hopefully this makes the role of the filter algorithms clear.
I felt that describing details of the pseudocode in words in the narrative would
be tedious and redundant and the pseudocode has comments that describe what is being
done whenever it is not obvious. Regarding figures, Section 3.4 in
page 52 has an example worked out in detail with figures. This was
added with the intention of illustrating the filter algorithms.

\item Comment: {\bf It is shown that TREE PATH LABELING PROBLEM on
    arbitrary trees can be solved polynomial time if the two
    subproblems, identified in section 3.6 in the thesis are
    polynomial time solvable. The formal definition of these two
    subproblems are not given. However, in page 71 item numbers 1 and
    2 the subproblems are described with several notations, where the
    motivation is unclear.}\\
Response: The prime submatrices partial order theory and notations are
required for a formal definition.
The two subproblems are defined now as \FOCPS and \FMFT and motivation is
given in Page 61 before setting up the theory and formal defintion is
given in page 66 after setting up the theory required to define it.

{\em

\begin{problemdef}{\FOCPS}{A hypergraph $\cF$ with its in-tree
    partitions $\{\bP_1,\ldots,\bP_r\}$ and
    a tree $T$.}
  Compute a partition of $T$ into subtrees $\{T_1, \ldots, T_r\}$ such
  that there exists an \ICPPL for $mub(\bP_i)$ from subtree
  $T_i$ for all $i \in [r]$ and there exists an ICPIA for all other
  prime submatrices in $\bP_i$ as claimed in Lemma 3.6.8.
\end{problemdef}

\begin{problemdef}{\FMFT}{An in-tree partition $\bP_i$ of a hypergraph
    $\cF$ and a subtree $T_i$ of a tree $T$.}
  Compute a feasible TPL for $mub(\bP_i)$ from subtree $T_i$.
\end{problemdef}
}

\item Comment: {\bf ... it is shown that consecutive 1s property
    testing for a \sout {graph} [hypergraph] is solvable in LOGSPACE. {\em This is a very
      important contribution.}}\\
Response: Since this is a conclusion from earlier results, it has not
been mentioned as a completely ``new'' result. It is mentioned in the abstract as a
derived result.
\end{enumerate}

### 2013-03-23 src/030chap3.tex
All changes made as per review. Viva presentation added.

In the remaining part of this section we show that $(\cF, \cl)$ is
feasible if and only if it is an ICPPL.
Algorithm~\ref{al:icppl-find-isomorph} takes $(\cF, \cl)$ as input and
computes the feasibility bijection if it is feasible or else reports
failure. This algorithm recursively does two levels of filtering of
$(\cF, \cl)$ to make it simpler in terms of set intersections while
retaining a non-empty subset of isomorphisms, if any, between $\cF$
and $\cF^\cl$. These two filters are \filteri in
Algorithm~\ref{al:filterone} and \filterii in
Algorithm~\ref{al:filtertwo}.
% One direction of this claim isclear: that if a path labeling is
% feasible, then all intersection cardinalities are preserved,
% i.e. the path labeling is an ICPPL. Algorithm~\ref{al:filterone}
% \annote{has no premature exit condition hence any input will go
%   through it}{Prove that the filtered sets has ICPPL \iff input PL
%   has ICPPL?}. Algorithm~\ref{al:filtertwo} has an exit condition at
% line~\ref{xempty}. It can be easily verified that $X$ cannot be
% empty if $\cl$ is a feasible path labeling. The reason is that a
% feasible path labeling has an associated bijection between
% $supp(\cF)$ and $V(T)$ \remove{i.e. $supp(\cF^{\cl})$} such that the
% sets map to paths, ``preserving'' the path labeling.  The rest of
% the section is devoted to constructively proving that it is
% sufficient for a path labeling to be an ICPPL and pass the two
% filtering algorithms.  To describe in brief, the constructive
% approaches refine an ICPPL iteratively, such that at the end of each
% iteration we have a ``filtered'' path labeling, and finally we have
% a path labeling that defines a family of bijections from $supp(\cF)$
% to $V(T)$\remove{ i.e. $supp(\cF^{\cl})$}.

@@

@@

%\noindent {\bf Finding an assignment of tree paths to a set system.} 
As described by Definition~\ref{def:matrixsetsystem} a set system or
hypergraph can be concisely represented by a binary matrix \stt the
row indices of the matrix denote the vertex set of the hypergraph and
each column bijectively correspond to a hyperedge. And thus the ideas
of feasible \TPL and \ICPPL are also defined for a binary matrix. If a
\TPL $(\cF,\cl,T)$ is feasible, \ie it is an ICPPL (see
Section~\ref{sec:feasible}) and if $M$ is the representative matrix
for hypergraph $\cF$, then we say that $M$ has an \ICPPL and vice
versa.

@@

\begin{lemma} 
  \label{lem:rooticppa}%
  Let $M$ be a matrix, $X$ be the directed graph whose vertices are in
  correspondence with the prime submatrices of $M$ and
  $\{\bP_1,\ldots,\bP_r\}$ be the partition of $X$ into in-trees by
  $\primeless$ defined earlier.  Let $T$ be the target tree and let
  $\{T_1, \ldots, T_r\}$ be a given partition of $T$ into vertex
  disjoint subtrees.  Then, for each $1 \leq i \leq r$, the set of
  prime submatrices $\bP_i$ has an ICPPL in $T_i$ if and only if the
  prime submatrix $mub(\bP_i)$ has an ICPPL on $T_i$ and all other
  matrices in $\bP_i$ have an ICPIA.
\end{lemma}
\begin{proof}\thesisspacing
  The proof is based on the fact that $\preccurlyeq$ is a partial
  order and $X$ is a directed graph which is a disjoint union of
  in-trees.  Each edge in the in-tree is a containment relationship
  among the supports of the corresponding submatrices. Therefore, any
  ICPPL to a prime submatrix that is not the $mub$ is contained in a
  path assigned to the sets in the parent matrix.  This follows from
  Lemma~\ref{lem:containment} and since $M_x \primeless mub(\bP_i)$
  for any prime submatrix $M_x \in \bP_i$.  Consequently, any ICPPL to
  the prime submatrix that is not $mub(\bP_i)$ is an ICPIA, and any
  ICPIA can be used to construct an ICPPL to the matrices
  corresponding to nodes in $\bP_i$ provided $mub(\bP_i)$ has
  an ICPPL in $T_i$. 
\end{proof}
Lemma \ref{lem:subicppa} and Lemma \ref{lem:rooticppa} point out two
algorithmic challenges in finding an ICPPL for a given hypergraph
$\cF$ on a target tree $T$.  
\begin{enumerate}
\item \label{it:challenge1} Finding $X$ and its partition into in-trees
  $\{\bP_1,\ldots,\bP_r\}$, can be done in polynomial time.  On the
  other hand, as per Lemma~\ref{lem:subicppa} we need to partition $T$
  into vertex disjoint subtrees $\{T_1, \ldots, T_r\}$ such that for
  each $i \in [r]$, the prime submatrices in $\bP_i$ have an ICPPL in
  $T_i$ (to be accurate, the prime submatrix $mub(\bP_i)$ has an ICPPL
  in $T_i$ and the rest of the prime submatrices have an ICPIA as
  claimed in Lemma~\ref{lem:rooticppa}).  This seems to be a
  challenging step, and it must be remarked that this step is easy
  when $T$ itself is a path, as each individual $T_i$ would be
  sub-paths.
\item \label{it:challenge2} The second algorithmic challenge is identified by
  Lemma~\ref{lem:rooticppa} which is to compute an ICPPL from the
  chosen subtree $T_i$
  to the matrix associated with $mub(\bP_i)$.
\end{enumerate}

>>>>

In the remaining part of this section we show that $(\cF, \cl)$ is
feasible if and only if it is an ICPPL.
Algorithm~\ref{al:icppl-find-isomorph} takes $(\cF, \cl)$ as input and
computes the feasibility bijection if it is feasible or else reports
failure. This algorithm recursively does two levels of filtering of
$(\cF, \cl)$ to make it simpler in terms of set intersections while
retaining a non-empty subset of isomorphisms, if any, between $\cF$
and $\cF^\cl$. These two filters are \filteri and \filterii which in
summary are as follows:
\begin{enumerate}
\item \filteri Refine $\cF$ such that the resulting
  labeling will not have paths that share a leaf thus each leaf being
  unique to a path. This is done by breaking the paths into subpaths and their
  corresponding preimage sets as described in Algorithm~\ref{al:filterone}.
\item \filterii Find the element in universe $U$ that maps to each
  leaf in $T$ as described in Algorithm~\ref{al:filtertwo}. 
\end{enumerate}
Remove the leaves from $T$ and their corresponding preimages from $U$
and call the filters again. This is repeated until the resulting
truncated tree is a path. The remaining mapping can be found using
ICPIA.

% One direction of this claim isclear: that if a path labeling is
% feasible, then all intersection cardinalities are preserved,
% i.e. the path labeling is an ICPPL. Algorithm~\ref{al:filterone}
% \annote{has no premature exit condition hence any input will go
%   through it}{Prove that the filtered sets has ICPPL \iff input PL
%   has ICPPL?}. Algorithm~\ref{al:filtertwo} has an exit condition at
% line~\ref{xempty}. It can be easily verified that $X$ cannot be
% empty if $\cl$ is a feasible path labeling. The reason is that a
% feasible path labeling has an associated bijection between
% $supp(\cF)$ and $V(T)$ \remove{i.e. $supp(\cF^{\cl})$} such that the
% sets map to paths, ``preserving'' the path labeling.  The rest of
% the section is devoted to constructively proving that it is
% sufficient for a path labeling to be an ICPPL and pass the two
% filtering algorithms.  To describe in brief, the constructive
% approaches refine an ICPPL iteratively, such that at the end of each
% iteration we have a ``filtered'' path labeling, and finally we have
% a path labeling that defines a family of bijections from $supp(\cF)$
% to $V(T)$\remove{ i.e. $supp(\cF^{\cl})$}.

@@

The two subproblems \FOCPS and \FMFT are informally defined as
follows. The precise definition is given after setting up the theory
and notations required in the following paragraphs.
\begin{enumerate}
\item \FOCPS - {\em Dividing the tree $T$ to subtrees for overlap
    component partitions: } Using the overlap partial ordering of
  hyperedges in $\cF$, the hyperedges can be divided into partitions
  of sub-hypergraphs (called partitions of prime submatrices) that are
  disjoint from each other. Being disjoint they must be assigned to
  disjoint subtrees of $T$ such that \FMFT can be solved successfully
  and the other sub-hypergraphs in the partition will have an ICPIA
  in accordance with the overlap property. Finding the right partition
  of $T$ into subtrees is the first subproblem.

\item \FMFT - {\em Finding ICPPL for $mub$ of each sub-hypergraph
    partition: } In each partition of sub-hypergraph, due to the
  overlap partial ordering, there will be an $mub$ sub-hypergraph that
  contains all the other sub-hypergraphs in that partition. If $T_i$
  is the subgraph assigned to this partition by solving \FOCPS, $T_i$
  is essentially the tree for the $mub$'s TPL. Computing an \ICPPL for
  this $mub$ sub-hypergraph from $T_i$ is the second subproblem.
\end{enumerate}

@@

%\noindent {\bf Finding an assignment of tree paths to a set system.}

As described by Definition~\ref{def:matrixsetsystem} a set system or
hypergraph can be concisely represented by a binary matrix \stt the
row indices of the matrix denote the vertex set of the hypergraph and
each column bijectively correspond to a hyperedge. And thus the ideas
of feasible \TPL and \ICPPL are also defined for a binary matrix. If a
\TPL $(\cF,\cl,T)$ is feasible, \ie it is an ICPPL (see
Section~\ref{sec:feasible}) and if $M$ is the representative matrix
for hypergraph $\cF$, then we say that $M$ has an \ICPPL and vice
versa.

@@

\begin{lemma} 
  \label{lem:rooticppa}%
  Let $M$ be a matrix, $X$ be the directed graph whose vertices are in
  correspondence with the prime submatrices of $M$ and
  $\{\bP_1,\ldots,\bP_r\}$ be the partition of $X$ into in-trees by
  $\primeless$ defined earlier.  Let $T$ be the target tree and let
  $\{T_1, \ldots, T_r\}$ be a given partition of $T$ into vertex
  disjoint subtrees.  Then, for each $1 \leq i \leq r$, the set of
  prime submatrices $\bP_i$ has an ICPPL in $T_i$ if and only if the
  prime submatrix $mub(\bP_i)$ has an ICPPL on $T_i$ and all other
  matrices in $\bP_i$ have an ICPIA.
\end{lemma}
\begin{proof}\thesisspacing
  The proof is based on the fact that $\preccurlyeq$ is a partial
  order and $X$ is a directed graph which is a disjoint union of
  in-trees.  Each edge in the in-tree is a containment relationship
  among the supports of the corresponding submatrices. Therefore, any
  ICPPL to a prime submatrix that is not the $mub$ is contained in a
  path assigned to the sets in the parent matrix.  This follows from
  Lemma~\ref{lem:containment} and since $M_x \primeless mub(\bP_i)$
  for any prime submatrix $M_x \in \bP_i$.  Consequently, any ICPPL to
  the prime submatrix that is not $mub(\bP_i)$ is an ICPIA, and any
  ICPIA can be used to construct an ICPPL to the matrices
  corresponding to nodes in $\bP_i$ provided $mub(\bP_i)$ has
  an ICPPL in $T_i$. 
\end{proof}
Lemma \ref{lem:subicppa} and Lemma \ref{lem:rooticppa} point out two
algorithmic challenges in finding an ICPPL for a given hypergraph
$\cF$ on a target tree $T$, namely subproblems \FOCPS and \FMFT respectively.

\item \label{it:challenge1} \FOCPS - Finding $X$ and its partition into
  in-trees $\{\bP_1,\ldots,\bP_r\}$, can be done in polynomial time.
  On the other hand, as per Lemma~\ref{lem:subicppa} we need to
  partition $T$ into vertex disjoint subtrees $\{T_1, \ldots, T_r\}$
  such that for each $i \in [r]$, the prime submatrices in $\bP_i$
  have an ICPPL in $T_i$ (to be accurate, the prime submatrix
  $mub(\bP_i)$ has an ICPPL in $T_i$ and the rest of the prime
  submatrices have an ICPIA as claimed in Lemma~\ref{lem:rooticppa}).
  This seems to be a challenging step, and it must be remarked that
  this step is easy when $T$ itself is a path, as each individual
  $T_i$ would be sub-paths.
\item \label{it:challenge2} \FMFT - The second algorithmic challenge is
  identified by Lemma~\ref{lem:rooticppa} which is to compute an ICPPL
  from the chosen subtree $T_i$ to the matrix associated with
  $mub(\bP_i)$.
\end{enumerate}

The formal definitions of the problems are given below.

\begin{problemdef}{\FOCPS}{A hypergraph $\cF$ with its in-tree
    partitions $\{\bP_1,\ldots,\bP_r\}$ and
    a tree $T$.}
  Compute a partition of $T$ into subtrees $\{T_1, \ldots, T_r\}$ such
  that there exists an \ICPPL for $mub(\bP_i)$ from subtree
  $T_i$ for all $i \in [r]$ and there exists an ICPIA for all other
  prime submatrices in $\bP_i$ as claimed in Lemma~\ref{lem:rooticppa}.
\end{problemdef}

\begin{problemdef}{\FMFT}{An in-tree partition $\bP_i$ of a hypergraph
    $\cF$ and a subtree $T_i$ of a tree $T$.}
  Compute a feasible TPL for $mub(\bP_i)$ from subtree $T_i$.
\end{problemdef}
